---
title: "Final Group Project"
authors: "Charry DeAndres - Kaniz Syeda - Shahnaz Jalali"
date: "11/17/2020"
dataset: "kc_house_data.csv"
output: html_document
---

# Setup libraries and Read King County housing dataset
```{r,warning=FALSE,message=FALSE}
#install.packages("leaflet", dependencies = TRUE)

# load required libraries
install.packages("tidyverse")
install.packages("GGally")
install.packages("caret", dependencies = c("Depends", "Suggests"))
install.packages("lattice")

library(tidyverse)
library(dplyr)
library(ggplot2)
library(GGally)
library(caret) 
library(corrplot)
library(stringr)
library(lubridate)
library(leaflet)

set.seed(1234)


# set working directory
setwd("C:/Users/Admin/Desktop/DATA")

# load king county housing dataset
houses = read.delim(file = "kc_house_data.csv", header = TRUE, sep = ",")

# get summary report
summary(houses)
```

# check for Missing Data
```{r,warning=FALSE,message=FALSE}
# display number of rows and first 6 rows of data
nrow(houses)
head(houses)

# check missing values: there was no NAs in the dataset
nrow(houses %>% na.omit()) == nrow(houses)
colSums(is.na(houses))
```

# Exploratory Data Analysis

```{r,warning=FALSE,message=FALSE}
# price distribution was skewed so we apply log() function
houses$log_price = log10(houses$price)

# Distribution of Price 
ggplot(houses, aes(x = price)) + 
  geom_density(fill = "blue") + 
  geom_histogram(binwidth = 1, fill = "blue") +
  ggtitle("Distribution of Price")

# Distribution of Log_Price
ggplot(houses, aes(x = log_price)) + 
  geom_density(fill = "blue") + 
  geom_histogram(binwidth = .10, fill = "blue") +
  ggtitle("Distribution of Log_Price")

# price per sqft living
houses$price_per_sqft_living = houses$price/houses$sqft_living
```

# Visualize the distribution of selected features
```{r, message=FALSE,warning=FALSE}
features <- c('bathrooms', 'sqft_living', 'sqft_above', 'sqft_lot', 'floors', 
              'sqft_basement', 'sqft_living15', 'zipcode','condition', 'grade')

for (i in (features)){
                        p <- ggplot(houses, aes_string(x = i)) +
                             geom_histogram() +
                             labs(title = str_c("Dustribution of ", i))
                      print(p)
                     }

# Distribution of yr_renovated 
houses %>% filter( yr_renovated != 0 ) %>%  
  ggplot(aes(x = yr_renovated)) + 
  geom_histogram(stat = "count") +
  ggtitle("Distribution of Year_Renovated")

# Distribution of yr_built
ggplot(houses, aes(x = yr_built)) + 
  geom_histogram(stat = "count") +
  ggtitle("Distribution of Year_Built")
``` 

# Features with correlation value of equal or greater than 40% 
```{r,warning=FALSE,message=FALSE}
# features with correlation value of equal or greater than 40% 
data_filtered = select(houses, -id, -lat, -long, -date, -zipcode, 
                       -yr_renovated, -price_per_sqft_living, -price )

column_names = colnames(data_filtered)

good_cor = function(name) {
                            cor_value = cor(data_filtered$log_price, data_filtered[name])
                            return(cor_value >= .40)
}

column_name_cor = Filter(good_cor, column_names)
data_filtered = select(data_filtered, column_name_cor)

# Plot filtered features with 40% or higher correlation to price
ggpairs(data_filtered) +
  ggtitle("Correlation of the features with 40% or higher correlation to log_price")
```

# House Price distribution comparing "with or without"" Waterfront, View, Basement or Renovation
```{r, message=FALSE,warning=FALSE}
# House price distribution comparing with or without "waterfront"
houses$waterfront = factor(houses$waterfront)
ggpairs(select(houses, log_price, waterfront), mapping = aes(color = waterfront, alpha = 0.5)) +
  ggtitle("House Price distribution comparing with or without waterfront")

# House Price distribution comparing with or without "view"
houses$has_view = ifelse(houses$view > 0, 1, 0)
houses$has_view = factor(houses$has_view)
ggpairs(select(houses, log_price, has_view), mapping = aes(color = has_view, alpha = 0.5)) +
  ggtitle("House Price distribution comparing with or without view")

# House Price distribution comparing with or without "basement"
houses$has_basement = ifelse(houses$sqft_basement > 0, 1, 0)
houses$has_basement = factor(houses$has_basement)
ggpairs(select(houses, log_price, has_basement), mapping = aes(color = has_basement, alpha = 0.5)) +
  ggtitle("House Price distribution comparing with or without basement")

# House Price distribution comparing with or without "renovation"
houses$has_renovation = ifelse(houses$yr_renovated > 0, 1, 0)
houses$has_renovation = factor(houses$has_renovation)
ggpairs(select(houses, log_price, has_renovation), mapping = aes(color = has_renovation, alpha = 0.5)) +
  ggtitle("House Price distribution comparing with or without renovation")
```

# House distribution of "sqft_lot" and "bedrooms" vs. price
```{r, message=FALSE,warning=FALSE}
# plot sqft_lot vs. price
ggplot(houses, aes(x = sqft_lot, y = log_price)) +
  geom_point() +
  geom_smooth(method = "lm") +
  ggtitle("Distribution of sqft_lot vs. Log_Price")

# plot bedrooms vs. price
ggplot(houses, aes(x = bedrooms, y = log_price)) +
  geom_point() +
  geom_smooth(method = "lm") +
  ggtitle("Distribution of Bedrooms vs. Log_Price")
```

## We’re curious how much is the Price per sq. ft of the living space, of the lot, and of the surrounding living spaces and lots. 
## Something happened in the King County Real Estate in Jan-Feb 2015. 
```{r, message=FALSE,warning=FALSE}
# collapses sales in one month to the first day of the month
houses$date <- paste(substr(houses$date, 1,6), "01", sep = "")
houses$date <- ymd(houses$date)

# calculates the price of square ft for each sqft related features
houses$price_sqft_living   = houses$price/houses$sqft_living
houses$price_sqft_lot      = houses$price/houses$sqft_lot
houses$price_sqft_lot15    = houses$price/houses$sqft_lot15
houses$price_sqft_living15 = houses$price/houses$sqft_living15

# computes the average price of sq ft 
houses <- houses %>% 
            mutate(price_sqft_avg = 
                (price_sqft_living + price_sqft_lot + price_sqft_lot15 + price_sqft_living15)/4)

# create data.frame "price_sqft_avg" by summarizing mean_price group by date
price_sqft_avg <- houses %>% 
                    group_by(date) %>% 
                    summarise(mean_price = median(price_sqft_avg)) %>% 
                    as.data.frame()

# Plot Average Price of Sq ft over Time
price_sqft_avg %>% ggplot(aes(x = date, y = mean_price)) +
                      geom_line(linetype = 2) +
                      geom_point(size = 2) +
                      ggtitle("Average Price of Sq ft over Time")

# Median price of houses for each Month over Time
median_price_by_month <- houses %>% 
                            group_by(date) %>% 
                            summarise(median_price_month = median(price))

# Plot Median Price over Time - Prices dip around Feb 2015.
median_price_by_month %>% 
  ggplot(aes(x = date, y = median_price_month)) +
    geom_line(linetype = 2) +
    geom_point(size = 2) +
    ggtitle("Median Price per Month over Time")

```

# Visualization of houses location using Esri Map 
```{r,warning=FALSE,message=FALSE}
houses$PriceBin <- cut(houses$price, c(0, 250e3, 500e3, 750e3, 1e6, 2e6, 999e6))
center_lon = median(houses$long, na.rm = TRUE)
center_lat = median(houses$lat, na.rm = TRUE)

factpal <- colorFactor(c("black", "blue", "yellow", "orange", "#0B5345", "red"), 
                       houses$PriceBin)

leaflet(houses) %>% addProviderTiles("Esri.NatGeoWorldMap") %>%
  addCircles(lng = ~long, lat = ~lat, 
             color = ~factpal(PriceBin))  %>%
  # controls
  setView(lng = center_lon, lat = center_lat, zoom = 12) %>%
  addLegend("bottomright", pal = factpal, values = ~PriceBin,
            title = "House Price Distribution",
            opacity = 1)
```

# Remove bedrooms outlier 
```{r}
# Houses with number of bedrooms <1 or >9
bedroom_outlier = houses %>% filter(houses$bedrooms < 1 | houses$bedrooms > 9 )
nrow(bedroom_outlier)
houses = houses %>% filter(between(bedrooms, 1, 9))
```

## Train/Test Dataset
```{r}
#Removing unnecessary features
houses_data = houses %>% select(-date, -price_per_sqft_living, -has_view, -has_basement,
                                -has_renovation, -price_sqft_living, -price_sqft_lot, -price_sqft_lot15,
                                -price_sqft_living15, -price_sqft_avg, -PriceBin, -price)

# split into train/test dataset
# 80% of data to be used as training
in_train = createDataPartition(y = houses_data$log_price, 
                               p = 0.8, 
                               list = FALSE)

houses_train = houses_data[in_train, ]
houses_train

houses_test =  houses_data[-in_train, ]
houses_test_example = test_house = data.frame(id = 0,
                                              bedrooms= 3,
                                              bathrooms= 1.75,
                                              sqft_living = 1520,
                                              sqft_lot = 6380,
                                              floors= 1,
                                              waterfront = factor(0),
                                              view= 0,
                                              condition= 3,
                                              grade = 7,
                                              sqft_above= 790,
                                              sqft_basement= 730,
                                              yr_built = 1948,
                                              yr_renovated=0,
                                              zipcode= 98115,
                                              lat= 47.6950,
                                              long= -122.304,
                                              sqft_living15= 1520,
                                              sqft_lot15= 6235,
                                              log_price= NA)

houses_test_example = bind_rows(houses_test,houses_test_example)

# Imbalanced features in data
nzv = nearZeroVar(houses_data, saveMetrics = TRUE)            
nzv %>% filter(nzv == TRUE)

# center/scale
preprocessing_steps = preProcess(select(houses_train, sqft_living, sqft_lot, sqft_above, 
                                        sqft_basement, sqft_living15, sqft_lot15), 
                                 method = c("center", "scale", "nzv"))

# apply pre-processing steps
houses_train_proc = predict(preprocessing_steps, newdata = houses_train)
houses_test_proc = predict(preprocessing_steps, newdata = houses_test)
houses_test_example_proc = predict(preprocessing_steps, newdata = houses_test_example)
```

# Model_1
```{r,warning=FALSE,message=FALSE}
# build full linear regresseion model using "RMSE" metrics and cross-validation of 10-fold
model_1 = train(log_price ~ ., 
                   data = houses_train_proc, 
                   method = 'lm', 
                   metric = 'RMSE',
                   tuneLength = 10,
                   trControl = trainControl(method = 'cv', number = 10))

# view model
model_1

# get summary report of the model_1
summary(model_1)
```

# Feature importance of Model_1
```{r,warning=FALSE,message=FALSE}
# plot feature importance of model_1
plot(varImp(model_1))
```

# Model_1 Prediction Metrics
```{r,warning=FALSE,message=FALSE}
# predict function using test data
pred_1 = predict(model_1, newdata = houses_test_proc)

# predicted target values
head(pred_1)

# calculate metrics by comparing prediction vs observation
postResample(pred = pred_1, obs = houses_test_proc$log_price)
```


```{r,warning=FALSE,message=FALSE}
# explicitly calculate the errors for each row
errors_1 = data.frame(predicted = pred_1,
                    observed = houses_test_proc$log_price,
                    error = pred_1 - houses_test_proc$log_price)

# plot the correlation between prediction and observation
ggplot(data = errors_1, aes(x = predicted, y = observed)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = 'red') +
  ggtitle("Model 1: Correlation between prediction and observation")
```

## Model_2
```{r,warning=FALSE,message=FALSE}
# build linear regresseion model using "RMSE" metrics and cross-validation of 10-fold 
model_2 = train(log_price ~ sqft_living + lat + grade + yr_built + view,
                   data = houses_train_proc,
                   method = 'lm', 
                   metric = 'RMSE',
                   tuneLength = 10,
                   trControl = trainControl(method = 'cv', number = 10))
# view model
model_2

# get summary report of the model
summary(model_2)
```

# Feature importance of model_2
```{r,warning=FALSE,message=FALSE}
# plot feature importance of model_2
plot(varImp(model_2))
```


# Model_2 Prediction Metrics
```{r,warning=FALSE,message=FALSE}
# predict function using test data
pred_2 = predict(model_2, newdata = houses_test_proc)

# predicted target values
head(pred_2)

# calculate metrics by comparing prediction vs observation
postResample(pred = pred_2, obs = houses_test_proc$log_price)
```

```{r,warning=FALSE,message=FALSE}
# explicitly calculate the errors for each row
errors_2 = data.frame(predicted = pred_2,
                    observed = houses_test_proc$log_price,
                    error = pred_2 - houses_test_proc$log_price)

# plot the correlation between prediction and observation
ggplot(data = errors_2, aes(x = predicted, y = observed)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = 'red') +
  ggtitle("Model 2: Correlation between prediction and observation")
```

# Model_3: Lasso Model
```{r,warning=FALSE,message=FALSE}
# build lasso model with cross-validation of 10-fold
model_3 = train(log_price ~ sqft_living + lat + grade + yr_built + view, 
                   data = houses_train_proc, 
                   method = 'lasso',
                   tuneLength = 10,
                   trControl = trainControl(method = 'cv', number = 10))

# view model
model_3

# get summary report of the model
summary(model_3)
```

# Feature importance of Lasso Model
```{r,warning=FALSE,message=FALSE}
# plot feature importance of lasso model
plot(varImp(model_3))
```

# Lasso Model Prediction Metrics
```{r,warning=FALSE,message=FALSE}
# predict function using test data
pred_3 = predict(model_3, newdata = houses_test_proc)

# predicted target values
head(pred_3)

# calculate metrics by comparing prediction vs observation
postResample(pred = pred_3, obs = houses_test_proc$log_price)
```


```{r,warning=FALSE,message=FALSE}
# explicitly calculate the errors for each row
errors_3 = data.frame(predicted = pred_3,
                    observed = houses_test_proc$log_price,
                    error = pred_3 - houses_test_proc$log_price)

# plot the correlation between prediction and observation
ggplot(data = errors_3, aes(x = predicted, y = observed)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = 'red') +
  ggtitle("Model 3: correlation between prediction and observation")
```

# Model Comparison
```{r,warning=FALSE,message=FALSE}
# compare the models using "resamples"
results = resamples(list(full_model = model_1, 
                         model_2 = model_2,
                         lasso_model = model_3))

# get summary of the results
summary(results)
```

# Plot the metrics
```{r,warning=FALSE,message=FALSE}
# plot the comparison metrics between the models
dotplot(results)
bwplot(results)
```


# Criteria of the house we are trying to buy or sell
```{r,warning=FALSE,message=FALSE}
example = houses_test_example_proc %>% filter(id == 0)
```

# Predict price on the criteria of our test house
```{r,warning=FALSE,message=FALSE}
predict(model_1, example)
```

# Convert log price to actual price
```{r,warning=FALSE,message=FALSE}
test_house_price = 10^ 5.634006       #replace this log_price with the current value from predict model result in line 453
test_house_price 
```
